Easy as Pi: A Network Coding Raspberry Pi Testbed

### 2016-06-01 schemes.tex
Expanded multicore description. 2 more citations added

The operations needed to encode and decode the data are, in general, similar. To
encode packets, the encoder needs to perform the matrix multiplication
$\textbf{C} = \textbf{V} \cdot \textbf{P}$. On the other hand, decoding the
information requires the decoder to find $\textbf{V}^{-1}$ and to perform the
matrix multiplication $\textbf{P} = \textbf{V}^{-1} \cdot \textbf{C}$. In both
cases, a matrix multiplication is needed. Therefore, to make a practical
implementation of network coding, it is valuable to find a way to optimize the
matrix multiplications operations for multicore architectures.

When designing multithreaded algorithms for network coding operations, it is
possible to implement the decoding by combining the matrix inversion and the
matrix multiplication, e.g., performing the Gauss-Jordan algorithm over the
coding matrix $\textbf{V}$ while performing, in parallel, row operations on the
coded data $\textbf{C}$. However, the parallelization of such operations provide
speeds up that are limited for small block sizes ($\leq 2048$ bytes). The reason
is that to operate in a parallel fashion over the same coded packet $C_i$
requires strained synchronization. In~\cite{5061951} and~\cite{4262451} the
parallelization of the row operations are optimized for \ac{GPU} and \ac{SMP}
systems respectively.

Therefore, to overcome the constraints of a tight synchronization, it is
preferable to explicitly invert the matrix $\textbf{V}$, and then take advantage
of optimizations for matrix multiplications. The authors implemented
in~\cite{wunderlich2015network} an algorithm that adopts the ideas of efficient
\ac{BLAS} operations~\cite{lawson1979basic} reimplementing them for finite
fields operations. The implemented algorithm tries to be cache efficient
maximizing the number of operations performed over a block of fetched data. This
is done by operating on square sub-blocks of the matrices, which exploits the
spatial locality of the data, at least for $\mathcal{O}(n^3)$
algorithms~\cite{golub2012matrix}. The optimal size of the block is
architecture-dependent. The ideal block has a number of operands that fit into
L1 cache, and it is a multiple of the \ac{SIMD} operations size.

The main idea of the implemented algorithm is to represent each one of the sub-
block matrix operations as a base, \textit{kernel}, operation that can be
optimized, e.g., using \ac{SIMD} operations. Each of these kernels, at the same
time, can be represented as a task with inputs and outputs in memory. If the
algorithm is formulated conventionally, and the data operations on the sub-
blocks are reordered, then it is possible to resolve data dependencies with a
\ac{DAG}. For the algorithm, a dependency means, for example, that a sub-block
of the matrix in memory cannot be read unless a previous writing operation was
performed there. When the \ac{DAG} is determined, then, a scheduler assigns to
the available cores the individual kernel operations that have their
dependencies fulfilled. The benefit of this algorithm is that the
synchronization relies only on data dependencies, and it does not requires the
insertion of artificial synchronization points. Using this technique, the matrix
inversion is performed using an algorithm based on LU
factorization~\cite{Dongarra:2011:HPM:2132876.2132885}, and the matrix
multiplication is performed by making the various matrix-matrix multiplications
of the sub-blocks.

>>>>

The operations needed to encode and decode data are, in general, similar. To
encode packets, the encoder needs to perform the matrix multiplication
$\textbf{C} = \textbf{V} \cdot \textbf{P}$. On the other hand, decoding the
information requires the decoder to find $\textbf{V}^{-1}$ and to perform the
matrix multiplication $\textbf{P} = \textbf{V}^{-1} \cdot \textbf{C}$. In both
cases, a matrix multiplication is needed. Therefore, to make a practical
implementation of network coding, it is valuable to find a way to optimize the
matrix multiplications operations for multicore architectures.

When designing multithreaded algorithms for network coding operations, it is
possible to implement the decoding by combining the matrix inversion and the
matrix multiplication, e.g., performing the Gauss-Jordan algorithm over the
coding matrix $\textbf{V}$ while performing, in parallel, row operations on the
coded data $\textbf{C}$. For example, in~\cite{5061951} and~\cite{4262451} the
parallelization of the row operations are optimized for \ac{GPU} and \ac{SMP}
systems respectively. However, the parallelization of such operations provide
limited speeds up for small block sizes ($\leq 2048$ bytes). The reason is that
to operate in a parallel fashion over the same coded packet $C_i$ requires
strained synchronization.

Therefore, to overcome the constraints of a tight synchronization, it is
preferable to explicitly invert the matrix $\textbf{V}$, and then take advantage
of optimizations for matrix multiplications, both at encoding and decoding time.
With that purpose, the authors implemented in~\cite{wunderlich2015network} an
algorithm that adopts the ideas of efficient \ac{BLAS}
operations~\cite{lawson1979basic} reimplementing them for finite fields
operations. Although there are already certain libraries available such as
~\cite{dumas2008dense} and \cite{dumas2002linbox} that allows the usage of the
highly optimize \ac{BLAS} implementations for some finite fields, they work
converting the GF elements into floating point numbers and back. Even though
that approach is efficient for large matrix sizes, the overhead imposed by the
conversions is not suitable for the the matrix sizes of typical network coding
implementations. The implemented algorithm aims to be cache efficient by
maximizing the number of operations performed over a block of fetched data. This
is done by dividing the matrices in square sub-blocks and operating with them.
As a consequence, this technique exploits the spatial locality of the data, at
least for $\mathcal{O}(n^3)$ algorithms~\cite{golub2012matrix}. The optimal size
of the block is architecture-dependent. The ideal block has a number of operands
that fit into L1 cache, and it is a multiple of the \ac{SIMD} operations size.

The main idea of the implemented algorithm is to represent each one of the sub-
block matrix operations (matrix-matrix multiplication, matrix-triangle matrix
multiplication, triangle-matrix system solving, etc) as a base, \textit{kernel},
operation that can be optimized individually, e.g., using \ac{SIMD} operations.
Each of these kernel operations, at the same time, can be represented as a task
with inputs and outputs in memory. If any algorithm is formulated
conventionally, and the data operations on the sub-blocks are reordered, then it
is possible to resolve data dependencies with a \ac{DAG}. For the algorithm, a
dependency means, for example, that a sub-block of the matrix in memory cannot
be read unless a previous writing operation was performed there. When the
\ac{DAG} is determined, then, a scheduler assigns to the available cores the
execution of the individual kernel operations that have their dependencies
satisfied. The benefit of this method is that the synchronization relies only on
data dependencies, and it does not requires the insertion of artificial
synchronization points. Using this technique, the matrix inversion is performed
using an algorithm based on LU
factorization~\cite{Dongarra:2011:HPM:2132876.2132885}, and the matrix
multiplication is performed by making the various matrix-matrix multiplications
on the sub-blocks.
